Reasoning about the rigor of perl programs: \\ \large gaps, limits to understanding, and means to ameliorate

### 2024-06-04 sciperl/paper.tex
largely good

    @misc{antithesis,
        title   = "Is something Bugging you?",
        author  = "Will Wilson",
        url     = {https://antithesis.com/blog/is_something_bugging_you/},
        year    = 2024,
        note    = "Having built similar systems myself, I second these sentiments."

@@

\subsection{I've got 100\% coverage though, its perfect right?}

It's usually fairly straightforward to get 100\% statement coverage on greenfield code.
However this rarely translates into software quality, as the test coverage on any of the dependencies in the stack is opaque in practically all programming languages.
How do we identify if the dependencies we use are well covered \textit{where we use them}?
If we can, the same techniques can be cross-referenced with the relative importance of the module vis. the "CPAN River" to identify where the most high-impact test coverage improvements could be made in the perl ecosystem.

It turns out that we can do this, supposing we actually run the testsuites of our dependencies when installing them.
This is (normally) the case when using something like perlbrew, with an isolated environment.
System perl is usually another story, as many modules will be shipped with another packaging manager.

While we could always wrap CPAN, ideally a CPAN::Plugin either pre or post test would be ideal.
Nevertheless, given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand, and cache it for later use, likely as a plugin to Devel::Cover instead.
cpancover \textit{appears} to do this, but is currently broken and appears unmaintained.

@@

Since not all programmers adhere to "Klingon Programming" (where it is better to \textit{die()} than \textit{return()} in failure),
there is no guarantee that environmental problems are caught.
This is why a great many Perl::Critic rules exist to catch precisely such problems caused by external factors.
For example:
\begin{lstlisting}
eval {...;1;} or do...

my $pid=fork(); die unless defined $pid;
\end{lstlisting}
Clearly, some means of actual fault injection should also be taken into account with regard to coverage.

How many subs are actually safe when `ls` returns `read error`?
Does your script recover from children whacked by oomkiller?
What if fork() fails because the process table is filled with zombies? \cite{gsb2}
What about when system() or qx{} or pipe open() where the target does not exist, but -f returns true because you have a bad RAID card?
I (and I suspect many of you) have had the misfortune of experiencing these cases personally.

This is actually a quite complicated subject.  As the founder of Antithesis puts it \cite{antithesis}:
\begin{displayquote}
the \$100 bill is surrounded by an entire field of poisonous spike-pits, nay, a continent of them.
\end{displayquote}
Aside from building something which can inject the \textit{astonishingly vast array} of such faults,
there is still the problem of integrating this into perl's testing harness, and reporting coverage of such in some fashion.

Given the scope, I have made no attempts at a proof of concept.
However, if I were to do so, I think the approach used by Test::MockFile \cite{mf} is likely one which can be leveraged to satisfy these needs.

@@

As mentioned earlier, the best means to evaluate whether changes to software harm it's fitness to purpose is some means of tracking user paths through the software as they perform tasks.
If sudden unintended disturbances in user paths are detected after pushing changes, it is a strong indication of error.
There are two levels of this; one is front-end and the other on the back end.
Ideally these are integrated somehow; given perl is not \textit{usually} the code powering UIs, but generating them, our primary concern will be with the points that interfaces reach out and touch backend code.

The best way to satisfy these needs is via a powerful logging framework using something like Log::Dispatch::DBI and robust session handling with shared identifiers between frontend and backend instrumentation.
From there you need something to \textit{identify and classify} common patterns of usage.
This sounds suspiciously like what Neural Networks are \textit{actually good at doing}.

AI::Categorizer running on a pack()ed representation of an application's state parameters throughout a user session ought to do the trick.
Yet another place I will have to rush to get a POC if I can at all.

@@

Devel::NYTProf is pretty much the best profiler in the industry; it's one of the primary reasons to recommend perl over alternatives.
However, the ability to reason about the accessibility, localizability or portability of your program is crude at best.
Generally the only option here is to impose linter rules which force the developer into the practices necessary to facilitate these things.
The only systematic attempt to apply linting across the perl ecosystem is CPANTS, but this could stand to be extended with more categories of evaluation.

For example, suppose I have the option of multiple modules on CPAN which I see do roughly the same thing in function X.
It would greatly simplify my decision as to what to use if I already knew ahead of time:
\begin{itemize}
\item What's the bandwidth of this function; how does it perform?
\item How complex is it?  I may have to read and debug this.
\item Is it doing non-portable things?
\end{itemize}
And so on.

This is also another place where NNs could be of great utility; while they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying it.
As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
You could roll right up to metacpan and see a "More like this" link -- like youtube but actually useful!
I for one would love seeing a table with these criteria and test coverage readily available -- no other packaging system does anything approaching this useful.

Oftentimes your choice of components is the primary determinant of your output's quality; we could do a lot more to help people reason about this.

@@

As you might expect from an active programming community of 35+ years, the things perl struggles to understand about its programs are essentially what every other programming language struggles with.
This should come as no shock, as the ultimate cause of them are pretty sticky fundamental issues of the "even people don't qualify as Strong AI" stripe.
The good news is that our near-total incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
As such the dream of "Bug Free" software is largely attainable in the same sense that a house in a bad neighborhood with cameras on it is going to be largely burgular free.
Being of relatively higher quality than all your competitors is highly desirable, as quality is the \textit{entire} driver of productivity and income.
You don't have to be original, use "alive" programming languages or even be particularly talented to win with this approach.

The software industry is one of the few places aside from SoEs where a 2/3 rate of failure \cite{standish} is remotely tolerable.
This is because our field is still relatively young (and thus high margin), but that is changing quickly.
The trend you need to get behind is not AI, but quality.
Whoever in our industry fully internalizes the lessons of Fred Brooks, Deming and Juran first will win.
Tooling such as suggested here would greatly further those ends, and give whatever language that builds them great advantage over their alternatives.

>>>>

@@

\subsection{100\% coverage, or: how many angels can dance upon the head of a pin?}

It's usually fairly straightforward to get 100\% statement coverage on greenfield\footnote{ex nihilo} code.
However this rarely translates into software quality, as the test coverage for any of the dependencies in the stack is opaque in practically all programming languages.
How do we identify if the dependencies we use are well covered \textit{where we use them}?

First, we must run the testsuites of our dependencies when installing them.
This is (normally) the case when using something like perlbrew, with an isolated environment.
System perl is usually another story, as many modules will be shipped with another packaging manager.

While we could always wrap CPAN, ideally a CPAN::Plugin either pre or post test would be ideal.
Given the wide variety of means that CPAN Modules might be installed, any solution here would have to build a coverage index on-demand (likely caching it for later use).
cpancover \textit{appears} it \textit{intends to} do this, but is currently broken and appears unmaintained.

The same techniques can be cross-referenced with the relative importance of the module.
This information is known as the \textit{CPAN River}\footnote{See http://neilb.org/2015/04/20/river-of-cpan.html}.
Doing this will identify where the most high-impact test coverage improvements can be made in the perl ecosystem.

@@

Since not all subroutines are \textit{Klingon Programming}\footnote{where it is better to \textit{die()} than \textit{return()} in failure.  Also known as "Control-Flow by Exception".  Such subroutines usually behave poorly in threaded contexts.},
such failures tend to go unnoticed when inured in this mental model.
This is why a great many Perl::Critic policies exist to catch precisely such problems caused by external factors largely outside of a subroutine's control.
For example, I have observed policies enforcing code along these lines:
\begin{lstlisting}
# Failing to have a 'do' clause after block eval without
# assignment generally means something is failing silently
eval {...;1;} or do {...}

# Failing to check that you got a defined pid from fork()
# means the program will have undefined behavior in the
# event the process table is full.
my $pid=fork() // die;
\end{lstlisting}
Some means of actual fault injection should also be taken into account in addition to standard coverage metrics as a measure of the code's overall robustness.

The sources of these faults outside of a subroutine's control are generally syscalls or FFI.
As such we will need some means of discerning whether or not the usage in a test behaves well under these scenarios:
\begin{itemize}
\item call emits (or is subjected to) a signal (e.g. SIGUSR1, SIGHUP, SIGKILL) either directly or in a child
\item call emits an exception
\item call returns a failure code
\item call appears successful, but nonetheless failed
\end{itemize}
Ideally, this would be done in a semi-automated fashion where one could write a test exercising the expected operational modes, and then re-execute subjected to permutations of these failures.
The approach used by \texttt{Test::MockFile}\cite{mock} is likely one which can be leveraged to inject syscall faults, and everything else can be handled with a universal mock (which would wrap whatever existing mock or sub is called.).

Automatically identifying all the subroutine calls within a given subroutine to inject these faults is complicated and relatively costly.
A look-ahead using a similar technique as used in \texttt{Devel::Trace}\footnote{https://metacpan.org/pod/Devel::Trace} (or internal interpreter knowledge) would need to be used to know whether we are in the right place.
Furthermore, we must pick the subroutines out of the next line if we are in the right place and wrap/mock them appropriately.

Thankfully there is a solution which uses XS to accomplish this: \texttt{Devel::Probe}\footnote{https://metacpan.org/pod/Devel::Probe}.

TODO: show off POC doing this with tests...

@@

As mentioned earlier, the best means to evaluate whether changes to software harm it's fitness to purpose is some means of tracking user paths through the software as they perform tasks.
If sudden unintended disturbances in user paths are detected after pushing changes, it is a strong indication of error.

This is generally done via coupling a powerful logging framework such as \texttt{Log::Dispatch::DBI}\footnote{https://metacpan.org/pod/Log::Dispatch::DBI} and robust session handling using unique identifiers.
From there you need something to \textit{identify and classify} common patterns of usage.
This sounds suspiciously like what Neural Networks are \textit{actually good at doing}.

\texttt{AI::Categorizer} running on a \texttt{pack()}ed representation of an application's state parameters throughout a user session ought to do the trick.

@@

Tools for reasoning about the performance of perl programs such as \texttt{Devel::NYTProf}\footnote{https://metacpan.org/pod/Devel::NYTProf} cover the subject of performance well.
However, the ability to reason about the accessibility, localizability or portability of your program is crude at best.
Generally the only option here is to impose linter rules which force the developer into the practices necessary to facilitate these things.
The only systematic attempt to apply such linting across the perl ecosystem is CPANTS\footnote{https://cpants.cpanauthors.org/},
but this could stand to be extended with more categories of evaluation.

For example, suppose I have the option of multiple modules on CPAN which I see do roughly the same thing in function X.
It would greatly simplify my decision as to what to use if I already knew ahead of time:
\begin{itemize}
\item how does it perform?
\item How complex is it?  I may have to read and debug this.
\item Is it doing non-portable things?
\end{itemize}
And so on.

This is also another place where AI techniques could be of great utility.
While they are necessarily awful at writing \textit{correct} code \cite{copilot}, they're pretty great at classifying code.
As such we could likely build these lists of "Modules that do largely the same things on CPAN" in a mostly automated fashion.
Currently, the best we have is the "SEE ALSO" section of a module's POD, which is far from comprehensive.

Coupling CPANTS to a classifier would make it far more useful, stimulating the development of these additional checks which do not currently exist.
Being able to sort a list of candidate modules for a particular purpose by \texttt{kwalitee}\footnote{CPANTS refers to non-functional parameters as kwalitee}
is one way to accomplish that.

@@

As you might expect from an active programming community of 35+ years, the things Perl programmers struggle to understand about their programs are what every programming language community struggles with.
The good news is that our incapacity to percieve all the ways in which our programs can fail also means automated solutions to ferreting these out are likely to be far more effective at doing so than users.
Much of this tooling has yet to have been built, and recent advances have put a number of them tantalizingly within our grasp.

The software industry is one of the few places aside from SoEs\footnote{State-Owned Enterprises} where a 2/3 rate of failure \cite{standish} is remotely tolerable.
This is because our field is still relatively young (and thus can afford such waste due to high profit margins) but that is changing quickly.
Programming communities that do not improve their tooling to enable writing programs with less defects will over time be out-competed by those that do.
